Unshuffling Permutations

### 2015-09-16 article.tex
Introduction hardness

We denote by $S_n$ the set of permutations of size $n$ and by $S$ the
set of all permutations. In this paper, permutations of a size $n$
are specified by words of length $n$ on the alphabet $[n]$ and without
multiple occurrence of a letter, so that all above definitions about
words remain valid on permutations. If $\pi$ is a permutation, we denote
by $\bar \pi$ (resp.~$\pi^{-1}$) the {\em complementary} (resp.
{\em inverse}) of $\pi$. If $u$ is a word of integers without multiple
occurrences of a same letter, $\STD(u)$ is the {\em standarized} of $u$,
that is the unique permutation of the same size as $u$ such that for all
$i, j \in [|u|]$, $u_i < u_j$ if and only if $\STD(u)_i < \STD(u)_j$. In
particular, the image of the map $\STD$ is the set $S$ of all
permutations. Two words $u$ and $v$ having the same standarized are
{\em order-isomorphic}. If $\sigma$ is a permutation, there is an
{\em occurrence} of (the {\em pattern}) $\sigma$ in $\pi$ if there is a
set $P$ of indexes of letters of $\pi$ such that $\sigma$ and $\pi_{|P}$
are order-isomorphic. When $\pi$ does not admit any occurrence of
$\sigma$, $\pi$ {\em avoids} $\sigma$.

@@

@@

\begin{lemma}
  \label{lemma:at most one edge monotone}
  Let $\pi = \pi_1 \, \sigma_1 \, \pi_2 \, \sigma_2 \, \pi_3$
  be a permutation with $|\sigma_1| \geq 2$ and $|\sigma_2| \geq 2$,
  and $\mathcal{M}$ be a containment-free perfect matching of $G_\pi$.
  The following assertions hold:
  \begin{enumerate}
    \item
    If $\sigma_1$ is increasing, $\sigma_2$ is decreasing and
    $\sigma_1$ is above $\sigma_2$
    (see Fig.~\ref{subfig:increasing before and aboce decreasing}),
    then $\mathcal{M}$ sontains at most one $(\sigma_1, \sigma_2)$-edge.
    \item
    If $\sigma_1$ is decreasing, $\sigma_2$ is increasing and
    $\sigma_1$ is below $\sigma_2$
    (see Fig.~\ref{subfig:decresing before and below increasing}),
    then $\mathcal{M}$ sontains at most one $(\sigma_1, \sigma_2)$-edge.
  \end{enumerate}
\end{lemma}

\begin{proof}[of Lemma~\ref{lemma:at most one edge monotone}]
  \begin{figure}[t!]
    \centering
    \subfigure[%
      An increasing pattern before and above a decreasing pattern.
    ]{%
      \begin{tikzpicture}
        [
          scale=0.45,
          label/.style={anchor=base}
        ]
        % increasing pattern
        \draw[step=1cm,black!30,ultra thin,fill=black!10] (-0.2,5.8) grid (5.2,11.2);
        \foreach \x/\y in {0/6,1/7,2/8,3/9,4/10,5/11} {
            \draw [fill=black] (\x,\y) circle (0.1);
        }
        % decreasing pattern
        \draw[step=1cm,black!30,ultra thin,fill=black!10] (5.8,-0.2) grid (11.2,5.2);
        \foreach \x/\y in {6/5,7/4,8/3,9/2,10/1,11/0} {
            \draw [fill=black] (\x,\y) circle (0.1);
        }
        % labels
        \node [label] (a) at (1,5) {$a$};
        \node [label] (b) at (4,5) {$b$};
        \node [label] (ap) at (7,-1) {$a'$};
        \node [label] (bp) at (10,-1) {$b'$};
        \node (nu1) at (2.5,4) {$\sigma_1$};
        \node (nu2) at (8.5,-2) {$\sigma_2$};
        % crossings edges
        \draw [line width=1pt]
        (1,7) .. controls +(0,12) and +(0,10) .. (7,4);
        \draw [line width=1pt]
        (4,10) .. controls +(0,12) and +(0,10) .. (10,1);
      \end{tikzpicture}
      \label{subfig:increasing before and aboce decreasing}
    }% end subfigure
    \qquad
    \subfigure[%
      A decreasing pattern before and below an increasing pattern.
    ]{%
    \begin{tikzpicture}
      [
        scale=0.45,
        label/.style={anchor=base}
      ]
      % decreasing pattern
      \draw[step=1cm,black!30,ultra thin,fill=black!10] (-0.2,-0.2) grid (5.2,5.2);
      \foreach \x/\y in {0/5,1/4,2/3,3/2,4/1,5/0} {
          \draw [fill=black] (\x,\y) circle (0.1);
      }
      % decreasing pattern
      \draw[step=1cm,black!30,ultra thin,fill=black!10] (5.8,5.8) grid (11.2,11.2);
      \foreach \x/\y in {6/6,7/7,8/8,9/9,10/10,11/11} {
          \draw [fill=black] (\x,\y) circle (0.1);
      }
      % labels
      \node [label] (a) at (1,-1) {$a$};
      \node [label] (b) at (4,-1) {$b$};
      \node [label] (ap) at (7,5) {$a'$};
      \node [label] (bp) at (10,5) {$b'$};
      \node (nu1) at (2.5,-2) {$\sigma_1$};
      \node (nu2) at (8.5,4) {$\sigma_2$};
      % crossings edges
      \draw [line width=1pt]
      (1,4) .. controls +(0,10) and +(0,12) .. (7,7);
      \draw [line width=1pt]
      (4,1) .. controls +(0,10) and +(0,12) .. (10,10);
      \end{tikzpicture}
      \label{subfig:decresing before and below increasing}
    }% end subfigure
    \caption{\label{fig:subfig:no (nu_1, nu_2)-edge}%
      Lemma~\ref{lemma:at most one edge monotone}:
      Assuming a $(\nu_1,\nu_2)$-edge $(a, b)$ in $\mathcal{M}$.
    }
  \end{figure}
  We first prove \emph{1}.
  Suppose, aiming at a contradiction, that there exist two
  $(\sigma_1, \sigma_2)$-edges $(a, a')$ and $(b, b')$ in $\mathcal{M}$.
  Since $\mathcal{M}$ is containment-free, we may assume $a < b$ and $a' < b'$.
  Therefore, we have
  $\pi(a) < \pi(b)$ (since $\sigma_1$ is increasing) and
  $\pi(a') > \pi(b')$ (since $\sigma_2$ is decreasing),
  which contradicts our hypothesis about $\mathcal{M}$.

@@

  For the sake of clarity, we will that a bunch of positions $P$
  is \emph{above} (resp. \emph{below}) above another bunch of positions $P'$
  in permutation $\mu$
  if $\mu(x) > \mu(y)$ (resp. $\mu(x) < \mu(y)$) for every $x \in P$
  and every $y \in P'$.
  For exemple, $\nu_1$ is above $\nu_2$ but below $\nu'_1$ in permutation $\mu$.

@@

      \node (claim edge 2) at (4,4) {Claim\ref{claim:(nu_2, nu'_2)-edge}};
      \node (claim edge 2) at (4,3.5) {Claim\ref{claim:all (nu_2, nu'_2)-edges}};

@@

      \node (claim edge 1) at (5.5,11.5) {Claim\ref{claim:(nu_3, nu'_3)-edge}};

@@

      \node (claim edge 1) at (8.5,8.5) {Claim\ref{}};

@@

      \node (claim edge 1) at (6.5,4) {Claim\ref{}};

@@

  We now turn to defining various gadgets (sequences of integers)
  that act as building blocks in our construction of a new permutation $\mu$:
  \begin{align*}
  \sigma'  &= ((k+1) \; \sigma \; (k+2)) \; [2N_2 + N_4 + 2n + k + 2] \\
  \pi'     &= ((n+1) \; \pi \; (n+2)) \; [2N_2 + N_4 + n + k + 2] \\
  \sigma'' &= \sigma \; [2N_2 + N_4] \\
  \pi''    &= \pi \; [2N_2 + N_4 + k] \\
  \nu_1    &= \nearrow_{N_1} \; [2N_2 + 2N_3 + 2N_4 + 2n + 2k + 4] \\
  \nu'_1   &= \nearrow_{N_1} \; [N_1 + 2N_2 + 2N_3 + 2N_4 + 2n + 2k + 4] \\
  \nu_2    &= \nearrow_{N_2} \; [N_2] \\
  \nu'_2   &= \searrow_{N_2} \\
  \nu_3    &= \nearrow_{N_3} \; [2N_2 + 2N_4 + 2n + 2k + 4] \\
  \nu'_3   &= \nearrow_{N_3} \; [2N_2 + N_3 + 2N_4 + 2n + 2k + 4] \\
  \nu_4    &= \searrow_{N_4} \; [2N_2 + N_4 + 2n + 2k + 4] \\
  \nu'_4   &= \searrow_{N_4} \; [2N_2]\text{.}
  \end{align*}
  We are now in position to define our target permutation $\mu$:
  $$
  \mu
  =
  \nu_1 \; \nu_2 \; \nu'_1 \; \nu_3 \; \sigma' \; \nu_4 \; \nu'_3 \; \pi' \; \nu'_4 \; \pi'' \; \sigma''
  \text{.}
  $$

  It is immediate that H can be constructed in polynomial-time in $n$ and $k$.
  We claim that $\sigma$ occurs as a pattern in $\pi$ if and only if
  there exists a containment-free perfect matching
  $\mathcal{M}$ in $G_\mu$ such that
  for any two disctinct edges
  $(a, b)$ and $(a', b')$ in $\mathcal{M}$
  we have $\mu(a) < \mu(a')$ if and only if $\mu(b) < \mu(b')$.

  Suppose first that $\sigma$ occurs in $\pi$ as a pattern.
  We show how to construct a containment-free perfect matching
  $\mathcal{M}$ in $G_\mu$ such that
  for any two disctinct edges
  $(a, b)$ and $(a', b')$ in $\mathcal{M}$
  we have $\mu(a) < \mu(a')$ if and only if $\mu(b) < \mu(b')$.
  \begin{itemize}
    \item $\mathcal{M}$ contains $N_1$ pairwise crossing
    $(\nu_1, \nu'_1)$-edges.
    \item $\mathcal{M}$ contains $N_2$ pairwise crossing
    $(\nu_2, \nu'_2)$-edges.
    \item $\mathcal{M}$ contains $N_3$ pairwise crossing
    $(\nu_3, \nu'_3)$-edges.
    \item $\mathcal{M}$ contains $N_4$ pairwise crossing
    $(\nu_4, \nu'_4)$-edges.
    \item $\mathcal{M}$ contains $k+2$ pairwise crossing
    $(\sigma', \pi')$-edges as depicted in Figure~\ref{}.
    (Notice that we use here the fact that $\sigma$ occurs as a pattern in $\pi$).
    \item $\mathcal{M}$ contains $n-k$ pairwise crossing
    $(\pi', \pi'')$-edges as depicted in Figure~\ref{}.
    \item $\mathcal{M}$ contains $k$ pairwise crossing
    $(\pi'', \sigma'')$-edges as depicted in Figure~\ref{}.
    (Notice that we use here again the fact that $\sigma$ occurs as a pattern in $\pi$).
  \end{itemize}
  It can be easily checked (probably referring to Figure~\ref{}) that
  $\mathcal{M}$ is perfect and that
  for any two disctinct edges
  $(a, b)$ and $(a', b')$ in $\mathcal{M}$
  we have $\mu(a) < \mu(a')$ if and only if $\mu(b) < \mu(b')$.

>>>>

We denote by $S_n$ the set of permutations of size $n$ and by $S$ the
set of all permutations. In this paper, permutations of a size $n$
are specified by words of length $n$ on the alphabet $[n]$ and without
multiple occurrence of a letter, so that all above definitions about
words remain valid on permutations. If $\pi$ is a permutation, we denote
by $\bar \pi$ (resp.~$\pi^{-1}$) the {\em complementary} (resp.
{\em inverse}) of $\pi$. If $u$ is a word of integers without multiple
occurrences of a same letter, $\STD(u)$ is the {\em standarized} of $u$,
that is the unique permutation of the same size as $u$ such that for all
$i, j \in [|u|]$, $u_i < u_j$ if and only if $\STD(u)_i < \STD(u)_j$. In
particular, the image of the map $\STD$ is the set $S$ of all
permutations. Two words $u$ and $v$ having the same standarized are
{\em order-isomorphic}. If $\sigma$ is a permutation, there is an
{\em occurrence} of (the {\em pattern}) $\sigma$ in $\pi$ if there is a
set $P$ of indexes of letters of $\pi$ such that $\sigma$ and $\pi_{|P}$
are order-isomorphic. When $\pi$ does not admit any occurrence of
$\sigma$, $\pi$ {\em avoids} $\sigma$.
We write $\nearrow_{n}$ (resp. $\searrow_{n}$)
for the permutation $1 2 \ldots n$ (resp. $n (n-1) \ldots 1$).
we write seqeunce of integers $s = s_1 s_2 \ldots s_n$, we
we let $s\;[k]$ stand for the lifted sequence
$(s_1+k) s_2+k) \ldots (s_n+k)$.

@@

Let $\pi$ be a permutation.
For the sake of clarity, we will say that a bunch of consecutive positions $P$
of $\pi$ is \emph{above} (resp. \emph{below}) above another bunch of consecutive
positions $P'$ in permutation $\pi$
if $\pi(i) > \pi(j))$ (resp. $\pi(i) < \pi(j)$) for every
$i \in P$ and every $j \in P'$.
For exemple, $\sigma_1$ is above $\sigma_2$
(in an equivalent manner, $\sigma_2$ is below $\sigma_1$)
in Fig.~\ref{subfig:increasing before and aboce decreasing},
whereas $\sigma_1$ is below $\sigma_2$
(in an equivalent manner, $\sigma_2$ is above $\sigma_1$)
in Fig.~\ref{subfig:decreasing before and below increasing}.

@@

\begin{lemma}
  \label{lemma:at most one edge monotone}
  Let $\pi = \pi_1 \, \sigma_1 \, \pi_2 \, \sigma_2 \, \pi_3$
  be a permutation with $|\sigma_1| \geq 2$ and $|\sigma_2| \geq 2$,
  and $\mathcal{M}$ be a containment-free perfect matching of $G_\pi$.
  The following assertions hold:
  \begin{enumerate}
    \item
    If $\sigma_1$ is increasing, $\sigma_2$ is decreasing and
    $\sigma_1$ is above $\sigma_2$
    (see Fig.~\ref{subfig:increasing before and aboce decreasing}),
    then $\mathcal{M}$ sontains at most one $(\sigma_1, \sigma_2)$-edge.
    \item
    If $\sigma_1$ is decreasing, $\sigma_2$ is increasing and
    $\sigma_1$ is below $\sigma_2$
    (see Fig.~\ref{subfig:decreasing before and below increasing}),
    then $\mathcal{M}$ sontains at most one $(\sigma_1, \sigma_2)$-edge.
  \end{enumerate}
\end{lemma}

\begin{proof}[of Lemma~\ref{lemma:at most one edge monotone}]
  \begin{figure}[t!]
    \centering
    \subfigure[%
      An increasing pattern before and above a decreasing pattern.
    ]{%
      \begin{tikzpicture}
        [
          scale=0.45,
          label/.style={anchor=base}
        ]
        % increasing pattern
        \draw[step=1cm,black!30,ultra thin,fill=black!10] (-0.2,5.8) grid (5.2,11.2);
        \foreach \x/\y in {0/6,1/7,2/8,3/9,4/10,5/11} {
            \draw [fill=black] (\x,\y) circle (0.1);
        }
        % decreasing pattern
        \draw[step=1cm,black!30,ultra thin,fill=black!10] (5.8,-0.2) grid (11.2,5.2);
        \foreach \x/\y in {6/5,7/4,8/3,9/2,10/1,11/0} {
            \draw [fill=black] (\x,\y) circle (0.1);
        }
        % labels
        \node [label] (a) at (1,5) {$a$};
        \node [label] (b) at (4,5) {$b$};
        \node [label] (ap) at (7,-1) {$a'$};
        \node [label] (bp) at (10,-1) {$b'$};
        \node (nu1) at (2.5,4) {$\sigma_1$};
        \node (nu2) at (8.5,-2) {$\sigma_2$};
        % crossings edges
        \draw [line width=1pt]
        (1,7) .. controls +(0,12) and +(0,10) .. (7,4);
        \draw [line width=1pt]
        (4,10) .. controls +(0,12) and +(0,10) .. (10,1);
      \end{tikzpicture}
      \label{subfig:increasing before and aboce decreasing}
    }% end subfigure
    \qquad
    \subfigure[%
      A decreasing pattern before and below an increasing pattern.
    ]{%
    \begin{tikzpicture}
      [
        scale=0.45,
        label/.style={anchor=base}
      ]
      % decreasing pattern
      \draw[step=1cm,black!30,ultra thin,fill=black!10] (-0.2,-0.2) grid (5.2,5.2);
      \foreach \x/\y in {0/5,1/4,2/3,3/2,4/1,5/0} {
          \draw [fill=black] (\x,\y) circle (0.1);
      }
      % decreasing pattern
      \draw[step=1cm,black!30,ultra thin,fill=black!10] (5.8,5.8) grid (11.2,11.2);
      \foreach \x/\y in {6/6,7/7,8/8,9/9,10/10,11/11} {
          \draw [fill=black] (\x,\y) circle (0.1);
      }
      % labels
      \node [label] (a) at (1,-1) {$a$};
      \node [label] (b) at (4,-1) {$b$};
      \node [label] (ap) at (7,5) {$a'$};
      \node [label] (bp) at (10,5) {$b'$};
      \node (nu1) at (2.5,-2) {$\sigma_1$};
      \node (nu2) at (8.5,4) {$\sigma_2$};
      % crossings edges
      \draw [line width=1pt]
      (1,4) .. controls +(0,10) and +(0,12) .. (7,7);
      \draw [line width=1pt]
      (4,1) .. controls +(0,10) and +(0,12) .. (10,10);
      \end{tikzpicture}
      \label{subfig:decreasing before and below increasing}
    }% end subfigure
    \caption{\label{fig:subfig:no (nu_1, nu_2)-edge}%
      Lemma~\ref{lemma:at most one edge monotone}:
      Assuming a $(\nu_1,\nu_2)$-edge $(a, b)$ in $\mathcal{M}$.
    }
  \end{figure}
  We first prove \emph{1}.
  Suppose, aiming at a contradiction, that there exist two
  $(\sigma_1, \sigma_2)$-edges $(a, a')$ and $(b, b')$ in $\mathcal{M}$.
  Since $\mathcal{M}$ is containment-free, we may assume $a < b$ and $a' < b'$.
  Therefore, we have
  $\pi(a) < \pi(b)$ (since $\sigma_1$ is increasing) and
  $\pi(a') > \pi(b')$ (since $\sigma_2$ is decreasing),
  which contradicts our hypothesis about $\mathcal{M}$.

@@

@@

      \node (claim edge 2) at (4,4) {Claim~\ref{claim:(nu_2, nu'_2)-edge}};
      \node (claim edge 2) at (4,3.5) {Claim~\ref{claim:all (nu_2, nu'_2)-edges}};

@@

      \node (claim edge 1) at (5.5,11.5) {Claim~\ref{claim:(nu_3, nu'_3)-edge}};

@@

      \node (claim edge 1) at (8.5,8.5) {Claim~\ref{}};

@@

      \node (claim edge 1) at (6.5,4) {Claim~\ref{}};

@@

  We now turn to defining various gadgets (sequences of integers)
  that act as building blocks in our construction of a new permutation $\mu$:
  \begin{align*}
  \sigma'  &= ((k+1) \; \sigma \; (k+2)) \; [2N_2 + N_4 + 2n + k + 2] \\
  \pi'     &= ((n+1) \; \pi \; (n+2)) \; [2N_2 + N_4 + n + k + 2] \\
  \sigma'' &= \sigma \; [2N_2 + N_4] \\
  \pi''    &= \pi \; [2N_2 + N_4 + k] \\
  \nu_1    &= \nearrow_{N_1} \; [2N_2 + 2N_3 + 2N_4 + 2n + 2k + 4] \\
  \nu'_1   &= \nearrow_{N_1} \; [N_1 + 2N_2 + 2N_3 + 2N_4 + 2n + 2k + 4] \\
  \nu_2    &= \nearrow_{N_2} \; [N_2] \\
  \nu'_2   &= \searrow_{N_2} \\
  \nu_3    &= \nearrow_{N_3} \; [2N_2 + 2N_4 + 2n + 2k + 4] \\
  \nu'_3   &= \nearrow_{N_3} \; [2N_2 + N_3 + 2N_4 + 2n + 2k + 4] \\
  \nu_4    &= \searrow_{N_4} \; [2N_2 + N_4 + 2n + 2k + 4] \\
  \nu'_4   &= \searrow_{N_4} \; [2N_2]\text{.}
  \end{align*}
  We are now in position to define our target permutation $\mu$
  (see Fig.~\ref{fig:reduction} for an illustration):
  $$
  \mu
  =
  \nu_1 \; \nu_2 \; \nu'_1 \; \nu_3 \; \sigma' \; \nu_4 \; \nu'_3 \; \pi' \; \nu'_4 \; \pi'' \; \sigma''
  \text{.}
  $$

  It is immediate that H can be constructed in polynomial-time in $n$ and $k$.
  We claim that $\sigma$ occurs in $\pi$ if and only if
  there exists a containment-free perfect matching
  $\mathcal{M}$ in $G_\mu$ such that
  for any two disctinct edges
  $(a, b)$ and $(a', b')$ in $\mathcal{M}$
  we have $\mu(a) < \mu(a')$ if and only if $\mu(b) < \mu(b')$.

  Suppose first that $\sigma$ occurs in $\pi$.
  We show how to construct a containment-free perfect matching
  $\mathcal{M}$ in $G_\mu$ such that
  for any two disctinct edges
  $(a, b)$ and $(a', b')$ in $\mathcal{M}$
  we have $\mu(a) < \mu(a')$ if and only if $\mu(b) < \mu(b')$.
  \begin{itemize}
    \item $\mathcal{M}$ contains $N_1$ pairwise crossing
    $(\nu_1, \nu'_1)$-edges.
    \item $\mathcal{M}$ contains $N_2$ pairwise crossing
    $(\nu_2, \nu'_2)$-edges.
    \item $\mathcal{M}$ contains $N_3$ pairwise crossing
    $(\nu_3, \nu'_3)$-edges.
    \item $\mathcal{M}$ contains $N_4$ pairwise crossing
    $(\nu_4, \nu'_4)$-edges.
    \item $\mathcal{M}$ contains $k+2$ pairwise crossing
    $(\sigma', \pi')$-edges as depicted in Figure~\ref{}.
    (Notice that we use here the fact that $\sigma$ occurs as a pattern in $\pi$).
    \item $\mathcal{M}$ contains $n-k$ pairwise crossing
    $(\pi', \pi'')$-edges as depicted in Figure~\ref{}.
    \item $\mathcal{M}$ contains $k$ pairwise crossing
    $(\pi'', \sigma'')$-edges as depicted in Figure~\ref{}.
    (Notice that we use here again the fact that $\sigma$ occurs as a pattern in $\pi$).
  \end{itemize}
  It can be easily checked (probably referring to Figure~\ref{}) that
  $\mathcal{M}$ is perfect and that
  for any two disctinct edges
  $(a, b)$ and $(a', b')$ in $\mathcal{M}$
  we have $\mu(a) < \mu(a')$ if and only if $\mu(b) < \mu(b')$.
